Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-14 analysis/paper/paper.qmd
updates of figures

::: {#fig-Spheroids-distribution-China}

@@

Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries.  1. Mujiaqiao, [@Wei1984]; 2. Liaohe river region, [@Li1991]; 3. Deshan Second Brick factory, [@Xi1994]; 4. Zhangjiatan, [@LixianMuseum1992]; 5. Huzhua Mount., [@Tan1999]; 6. Fangniu Mount., [@Fang2002]; 7. Fanba, [@Xia2022Fanba]; 8. Xiaokongshan, [@XiaokongshanTeam1988]; 9. Hejialiang, [@Wang2014Hanzhong]; 10. Longgangsi, [@Lu2006Liangshan]; 11. Shangdan basin, [@Wang2013Danjiang]; 12. Yaoshi basin, [@Wang2000Yaoshi]; 13. Choushuihe, [@Dai1973Lantian]; 14. Kehe, [@Jia1962Kehe]; 15. Shuigou, [@Huang1964Sanmenxia]; 16. Hezhigou, [@Liu1984Jingchuan]; 17. Xigou, [@Jia1959Xigou]; 18. Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19. Liujiacha, [@Xie1982Liujiacha]; 20. Dangcheng, [@Wu1989Dangcheng] ; 21. Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22. Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23. Jinsitai, [@Wang2010Jinstai]

:::

@@

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Cores",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )) %>% 
         filter(category != "Total"), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#66c2a5",
    "Cores" = "lightgray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.87, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black"),
    legend.box.margin = margin(0, 10, 0, 10)
  )

@@

# 画图
p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Cores" = "lightgray", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#66c2a5"
  )) +
  theme_minimal(base_size = 8) +
  guides(fill = "none")

@@

plot_grid(p_raw_material,
          p_dimensions,
          labels = "AUTO")

@@

# Define the position for the text block
x_pos <- 0.875
y_pos_start <- 1.04
v_spacing <- 0.005 # The vertical gap between lines

@@

# 画图
p_sphericity_and_she <- 
ggplot(df_long, aes(x = Group, 
                    y = Value, 
                    fill = Group,
                    colour = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, 
              alpha = 1, 
              size = 1) +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Spheroid\nSubspheroid" = "orange", 
                               "Polyhedron\nMultifacial" = "#66c2a5")) +
  scale_colour_manual(values = c("Spheroid\nSubspheroid" = "orange", 
                               "Polyhedron\nMultifacial" = "#66c2a5")) +
  labs(x = "",
       y = "",
       fill = "Group", 
      # title = "Sphericity and SHE by Group"
       ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none"
  )
```

@@

# 绘图，带误差带
p_power_spectrum <- 
ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  labs(
   # title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
  legend.position = c(0.95, 0.4),
  legend.justification = c("right", "top"),
  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
  legend.box.background = element_rect(color = "gray80"),
  legend.box.margin = margin(5, 5, 5, 5)
  )
```

@@

# Subset for labeling
label_data <- umap_data %>%
  filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
```

```{r umap_by_category, include=FALSE}
category_colors <- c(
  "idealmodel" = "black",
  "spheroid" = "red",
  "subspheroid" = "orange",
  "multifacial" = "lightgray",
  "polyhedron" = "gray40"
)

@@

```{r, echo=FALSE}
#| label: fig-UMAP-SPHARM-clusters
#| fig-cap: "UMAP plot showing distinct clusters of shapes derived from SPHARM"

@@

```{r, echo=FALSE}
#| label: fig-UMAP-SPHARM-samples
#| fig-cap: "UMAP plot of SPHARM-derived shape features colored by types"

@@

@@

ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
  geom_mask() + 
  geom_point(size = 3) + 
  theme_bw() +
  labs(
    x = "Planar", 
    y = "Isotropic", 
    yarrow = "Isotropy Ratio", 
    z = "Linear", 
    zarrow = "Elongation Ratio"
  ) +
  theme_showarrows() + 
  theme_clockwise() + 
  #geom_text(position = pn, aes(label = type), check_overlap = FALSE, size = 3) +
  scale_color_manual(values = c(
    "multifacial" = "lightgray",
    "polyhedron" = "gray40",
    "subspheroid" = "orange",
    "spheroid" = "red"
  ))

@@

According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids from Qianshangying. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@@fig-basic-characteristics). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

@@

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids. <!-- BM: check Cabanès,  do we agree or disagree with their findings-->

@@

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids. One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.

>>>>

::: {#fig-Spheroids-distribution-China}
```{r, echo=FALSE}

@@

Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries. 1. Mujiaqiao, [@Wei1984]; 2. Liaohe river region, [@Li1991]; 3. Deshan Second Brick factory, [@Xi1994]; 4. Zhangjiatan, [@LixianMuseum1992]; 5. Huzhua Mount., [@Tan1999]; 6. Fangniu Mount., [@Fang2002]; 7. Fanba, [@Xia2022Fanba]; 8. Xiaokongshan, [@XiaokongshanTeam1988]; 9. Hejialiang, [@Wang2014Hanzhong]; 10. Longgangsi, [@Lu2006Liangshan]; 11. Shangdan basin, [@Wang2013Danjiang]; 12. Yaoshi basin, [@Wang2000Yaoshi]; 13. Choushuihe, [@Dai1973Lantian]; 14. Kehe, [@Jia1962Kehe]; 15. Shuigou, [@Huang1964Sanmenxia]; 16. Hezhigou, [@Liu1984Jingchuan]; 17. Xigou, [@Jia1959Xigou]; 18. Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19. Liujiacha, [@Xie1982Liujiacha]; 20. Dangcheng, [@Wu1989Dangcheng] ; 21. Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22. Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23. Jinsitai, [@Wang2010Jinstai]
:::

@@

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Cores",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )) %>% 
         filter(category != "Total"), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#b2abd2",
    "Cores" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.87, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black"),
    legend.box.margin = margin(0, 10, 0, 10)
  )

@@

# 画图
p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Cores" = "#b2abd2", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  guides(fill = "none")

@@

p_combined <- plot_grid(
  p_raw_material,
  p_dimensions,
  labels = "AUTO",
  ncol = 2
)

# 再读取回来显示
ggsave(
  filename = here("analysis/paper/figures/basic_characteristics.png"),
  plot = p_combined,
  width = 9,   
  height = 4,  
  dpi = 300
)

knitr::include_graphics(here("analysis/paper/figures/basic_characteristics.png"))
```

@@

# Define the position for the text block
x_pos <- 0.875
y_pos_start <- 1.06
v_spacing <- 0.005 # The vertical gap between lines

@@

# 画图
p_sphericity_and_she <- 
ggplot(df_long, aes(x = Group, 
                    y = Value, 
                    fill = Group,
                    colour = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, 
              alpha = 1, 
              size = 1) +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Spheroid\nSubspheroid" = "orange", 
                               "Polyhedron\nMultifacial" = "#b2abd2")) +
  scale_colour_manual(values = c("Spheroid\nSubspheroid" = "orange", 
                               "Polyhedron\nMultifacial" = "#b2abd2")) +
  labs(x = "",
       y = "",
       fill = "Group", 
      # title = "Sphericity and SHE by Group"
       ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none"
  )
```

@@

# 绘图，带误差带
p_power_spectrum <- 
ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "grey60",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#b2abd2"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "grey60",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#b2abd2"
  )) +
  labs(
   # title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
  legend.position = c(0.95, 0.4),
  legend.justification = c("right", "top"),
  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
  legend.box.background = element_rect(color = "gray80"),
  legend.box.margin = margin(5, 5, 5, 5)
  )
```

@@

# Subset for labeling
label_data <- umap_data %>%
  filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))

```{r umap_by_category, include=FALSE}
category_colors <- c(
  "idealmodel" = "black",
  "spheroid" = "#e66101",
  "subspheroid" = "#fdb863",
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99"
)

@@

```{r, include=FALSE}
#| label: fig-UMAP-SPHARM-clusters
#| fig-cap: "UMAP plot showing distinct clusters of shapes derived from SPHARM"

@@

```{r, include=FALSE}
#| label: fig-UMAP-SPHARM-samples
#| fig-cap: "UMAP plot of SPHARM-derived shape features colored by types"

@@

```{r}
#| label: fig-UMAP-combined
#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
#| echo: false

library(magick)    
library(cowplot)

p_panel_spharm <- plot_grid(
  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_clusters.png")),
  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_samples.png")),
  labels = "AUTO",
  ncol = 1
)

ggsave(
  plot = p_panel_spharm,
  filename = here("analysis/paper/figures/UMAP_combined.png"),
  width = 9,
  height = 10,
  dpi = 300
)

knitr::include_graphics(here("analysis/paper/figures/UMAP_combined.png"))

@@

ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
  geom_mask() + 
  geom_point(size = 3) + 
  theme_bw() +
  labs(
    x = "Planar", 
    y = "Isotropic", 
    yarrow = "Isotropy Ratio", 
    z = "Linear", 
    zarrow = "Elongation Ratio"
  ) +
  theme_showarrows() + 
  theme_clockwise() + 
  #geom_text(position = pn, aes(label = type), check_overlap = FALSE, size = 3) +
  scale_color_manual(values = c(
    "multifacial" = "#b2abd2",
    "polyhedron" = "#5e3c99",
    "subspheroid" = "#fdb863",
    "spheroid" = "#e66101"
  ))

@@

According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids from Qianshangying. This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (\@@fig-basic-characteristics). This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.

@@

These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

@@

However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores. In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns. At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable. As @cabanès2022 mentioned, they could share a common operative chain before completing the final product. However, it remains unclear whether this typological transformation was intentional or unplanned. In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids. One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
